---
title: "Working with Centered Coverage Matrices"
author: "Tamas Schauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Centered Coverage Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* ChIP-seq data using H2Av antibody in Drosophila
* Coverages are centered at transcription start sites (TSS)
* See code to generate them at the composite plot documentation
* Subset, plot and summarize such matrices

## Setup Data

```{r message=FALSE, warning=FALSE,results=FALSE}
library(HelpersforChIPSeq)
```

```{r}
data_dir <- system.file("extdata/", package = "HelpersforChIPSeq")

my_mats <- file.path(data_dir ,list.files(path = data_dir, pattern = "^mat"))

for(i in seq_along(my_mats)){
        
        load(my_mats[i])
}

my_mats <- ls(pattern = "^mat")
my_mats
 
# Note: these matrices were binned with a size of 10
# See composite plot tutorial how to generate matrices
my_binning <- 10

# get one of the matrices as example
my_mat <- get(my_mats[4])

# matrix dimensions correspond to TSS +/- 2000 bp
dim(my_mat)

# example rows and columns
my_mat[1:5,201:210] 
```


## Row of the matrix

```{r message=FALSE, warning=FALSE,resultsTRUE, fig.align='center', fig.height=7, fig.width=7, dpi=200, out.width='75%'}
# select example row
my_mat_row <- my_mat[rownames(my_mat) == "FBgn0031282",]

length(my_mat_row)

# plot example row
plot(my_mat_row, type="l")

# draw center line
abline(v = (length(my_mat_row)+1)/2, lty=2)
```

```{r message=FALSE, warning=FALSE,results=TRUE, fig.align='center', fig.height=7, fig.width=7, dpi=200, out.width='75%'}
# smooth example row
my_mat_row_smooth <- zoo::rollmean(my_mat_row, 11)

# note length shortening
length(my_mat_row_smooth)

plot(my_mat_row_smooth, type="l")

# new center
abline(v = (length(my_mat_row_smooth)+1)/2, lty=2)
```

## Filtering

```{r message=FALSE}
# filter based on row standard deviation
# "flat" rows get removed
library(matrixStats)

my_filter <- rowSds(my_mat) > 0.25
my_mat <- my_mat[my_filter,]

dim(my_mat)
```

## Average matrix


```{r message=FALSE, warning=FALSE,results=FALSE, fig.align='center', fig.height=7, fig.width=7, dpi=200, out.width='75%'}
# plot average matrix
plot(colMeans(my_mat), type="l")

# subset central 1000 bp
plot(colMeans(my_mat[,151:250]), type="l")
```



## Ordering / sorting

```{r message=FALSE}
# take row average of a subset (500 bp downstream of the center)
row_ave <- rowMeans(my_mat[,201:250])

# take decreasing order of the averages 
my_order <- order(row_ave, decreasing = T)

# sort/order matrix based on these averages
my_mat <- my_mat[my_order,]

# new order
my_mat[1:5,201:210]
```

## Simple heatmap

```{r message=FALSE, warning=FALSE,results=FALSE, fig.align='center', fig.height=7, fig.width=4, dpi=200, out.width='50%'}
# image of the matrix
# Note matrix is transposed and flipped
image(my_mat, 
      useRaster = TRUE)

# correct for proper orientation
image(t(my_mat)[,nrow(my_mat):1], 
      useRaster = TRUE)

# set color palette
library(RColorBrewer)
my_colors <- colorRampPalette(c("white","orange","red","black"))(100)

# set contrast with min and max values
# Note: do not do calculations on such modified matrix
min_value <- 1
max_value <- 7

my_mat[my_mat < min_value] <- min_value
my_mat[my_mat > max_value] <- max_value

# set color breaks (steps) based on set values
my_breaks <- seq(min(min_value), max(max_value), length.out = 101)

image(t(my_mat)[,nrow(my_mat):1],
      col = my_colors,
      breaks =  my_breaks,
      useRaster = TRUE)
```


```{r message=FALSE, warning=FALSE,results=FALSE, fig.align='center', fig.height=2, fig.width=4, dpi=200, out.width='50%'}
# generate own color key
image(matrix(seq_along(my_colors)), 
      col = my_colors, 
      axes=FALSE, useRaster = TRUE)

axis(side = 1, at = c(0,0.5,1), 
     labels = round(my_breaks[c(1,51,101)],1))
```

* All this code was wrapped in helper functions 
* See heatmap documentation


```{r}
sessionInfo()
```
